### 2019-08-08 reports/thesis/chapters/3_compoundPoisson.tex
inference chapter
subroi for more projections
use of dotsc
plot sub roi section

The compound Poisson distribution is studied here because of the compound Poission-like behaviour from the detection of photons \citep{whiting2002signal, elbakri2003efficient, whiting2006properties}. It is defined by defining a latent variable $Y\sim\poisson(\lambda)$ with probability mass function (p.m.f.) $\prob(Y=y)=\euler^{-\lambda}\frac{\lambda^y}{y!}$ for $y=0,1,2,\cdots$, where $\lambda>0$ is the Poisson rate parameter. Let $U_i$ be some independent and identically distributed (i.i.d.) latent random variables with probability density function (p.d.f.) $p_U(u)$ for $i=1,2,3,\cdots$. Let $X$ be a compound Poisson random variable where
\begin{equation}
  X|Y = \sum_{i=1}^{Y}U_i \ .
  \label{eq:compoundPoisson_X|Y}
\end{equation}
The p.d.f.~of $X$ can be obtained by marginalising the joint p.d.f.
\begin{equation}
  p_X(x)=\sum_{y=0}^\infty p_{X|Y}(x|y)\prob(Y=y) \quad\text{for }x\geqslant 0
  \ .
\end{equation}
It should be noted that $X=0$ if and only if $Y=0$ and this happens with probability $\prob(Y=0)=\euler^{-\lambda}$. This implies that $X$ has probability mass at zero and probability density for positive numbers which results in the p.d.f.
\begin{equation}
  p_X(x) = 
  \begin{cases}
    \delta(x) \euler^{-\lambda}  & \text{ for } x=0 \\ 
    \sum_{y=1}^\infty p_{X|Y}(x|y)\euler^{-\lambda}\frac{\lambda^y}{y!} \quad\text{for } & \text{ for } x>0
  \end{cases}
\end{equation}
where $\delta(x)$ is the Dirac delta function.

@@

In an x-ray tube, photons are emitted as a Poisson process and each photon has some random energy due to bremsstrahlung and characteristic radiation. This shares similarities to the compound Poisson. Let $Y$ be the number of photons emitted for some time exposure $\tau$, then $Y\sim\poisson(\lambda)$. Each photon is assumed to be i.i.d.~with energy $U_i$ for $i=1,2,3,\cdots$. $U_i$ has p.d.f.~$p_U(u)$. The random variables discussed here covers all the latent variables in the compound Poisson.

@@

The evaluation of the p.d.f.~using the saddlepoint approximate required a bit of caution to avoid over/underflow problems. Suppose realizations of $X$ were simulated $\left\{x_1,x_2,x_3,\cdots,x_n\right\}$. The saddlepoint approximation was computed up to a constant using
\begin{equation}
  p_X(x) \propto
  \exp\left[
    -\frac{\alpha+2}{2(\alpha+1)}
    \ln(x)
    -x\beta
    +\left(
    \frac{x\beta}{\alpha}
    \right)^{\frac{\alpha}{\alpha+1}}\lambda^{\frac{1}{\alpha+1}}(\alpha+1) - k
  \right]
\end{equation}
for 500 equally spaced points from and including the minimum non-zero simulated value to the maximum simulated value. $k$ is some constant which was chosen to be
\begin{equation}
  k =
  \max_{i\in\left\{1,2,3,\cdots,n\right\}}
  \left[
    -\frac{\alpha+2}{2(\alpha+1)}
    \ln(x_i)
    -x_i\beta
    +\left(
      \frac{x_i\beta}{\alpha}
    \right)^{\frac{\alpha}{\alpha+1}}\lambda^{\frac{1}{\alpha+1}}(\alpha+1)
  \right]
  \ .
\end{equation}
The density was then was normalised by numerically integrating it using the trapezium rule using the 500 evaluated points.

A Q-Q plot is a plot which compares the empirical quantiles with the theoretical quantiles. Let
\begin{equation}
  F_X(x) = \prob(X\leqslant x)
\end{equation}
and
\begin{equation}
  \widehat{F}_X(x) = \frac{1}{n}\cdot\text{max}
  \left[
    \left(\sum_{i=1}^n\mathbb{I}(x_i\leqslant x)\right)-0.5,0
  \right]
  \ ,
\end{equation}
then a Q-Q plot is a parametric plot which plots $\widehat{F}_X^{-1}(p)$ against $F_X^{-1}(p)$ for $p=\frac{0.5}{n},\frac{1.5}{n},\frac{2.5}{n},\cdots,\frac{n-0.5}{n}$. If $F_X(x)$ and $\widehat{F}_X(x)$ are similar, a Q-Q plot should be a straight line with gradient 1, intercepting the origin. For the exact method and the saddlepoint approximate, $F_X(x)$ was found numerically by evaluating the p.d.f.~at $10\,000$ equally spaced points from and including the minimum to the maximum of the simulated samples, and then summing the required trapeziums. $\widehat{F}_X^{-1}(p)$ was then calculated by interpolation. Because the Normal and saddlepoint approximation does not support zero, the Q-Q plot was omitted for small $\lambda$.

@@

The EM algorithm \citep{dempster1977maximum} is proposed as a method to estimate the parameters of a $X\sim\CPoisson(\lambda,\alpha,\beta)$ random variable given a sample of measurements of it $\left\{x_1,x_2,x_3,\cdots,x_n\right\}$. The use of the EM algorithm here for the compound Poisson is similar to \cite{elbakri2003statistical, xu2009electronic}.

@@

The EM algorithm \citep{dempster1977maximum} instead treats $Y\sim\poisson(\lambda)$ as a latent variable. Let $\left\{Y_1,Y_2,Y_3,\cdots, Y_n\right\}$ be realizations of $Y$ and $\left\{x_1,x_2,x_3,\cdots, x_n\right\}$ be the respective measurements of $X|Y$. Define the joint log likelihood to be
\begin{multline*}
  \ln L(\lambda,\alpha,\beta;X,Y)
  =
  \sum_{i=1}^n
  \left[
    \mathbb{I}(x_i=0)
    \ln
    \prob(Y=0)
  \right.
  \\
  \left.
    +
    \mathbb{I}(x_i>0)
    \ln
    \left[
      p_{X|Y}(x_i|Y_i)\prob(Y=Y_i)
    \right]
  \right]
\end{multline*}
so that
\begin{multline}
  \ln L(\lambda,\alpha,\beta;X,Y)=
  \sum_{i=1}^n
  \left[
    -\mathbb{I}(x_i=0)
    \lambda
  \right.
  \\
  \left.+
    \mathbb{I}(x_i>0)
    \left(
      Y_i\alpha\ln\beta-\ln\Gamma(Y_i\alpha)+(Y_i\alpha-1)\ln x_i - \beta x_i
    \right.
  \right.
  \\
  \left.
    \left.  
      - \lambda + Y_i \ln \lambda - \ln(Y_i!)
    \right)
  \right]
  \ .
\end{multline}
The estimators are found by optimising the joint log likelihood. This is done iteratively by estimating the $Y$s given the $X$s and parameters (E step), followed by estimating the parameters given the $Y$s and $X$s (M step) until some convergence conditions are met.

@@

Let
\begin{equation}
  W_y^{(r)} = y^r W_y \quad \text{for }r=1,2,3,\cdots
\end{equation}
so that
\begin{equation}
  y_i = \frac{\sum_{y=1}^\infty W^{(1)}_y}{\sum_{y=1}^\infty W_y} \ .
\end{equation}
Similarly
\begin{equation}
  \zeta_i = \variance[Y|X=x_i] = \frac{\sum_{y=1}^\infty W^{(2)}_y}{\sum_{y=1}^\infty W_y} - \left(y_i\right)^2
  \ .
\end{equation}
The expectation terms, $y_i$ and $\zeta_i$, are evaluated in the M step so that they can be used in the E step. The evaluation can be done by truncating the sum
\begin{equation}
  \sum_{y=1}^\infty W^{(r)}_y \approx \sum_{y=y_\text{l}}^{y_\text{u}} W^{(r)}_y
\end{equation}
where $y_\text{l}<y_\text{max}<y_\text{u}$ and $y_\text{max}$ is the value of $y$ which maximises $W_y^{(r)}$. By expressing $\ln(W^{(r)}_y)$ as
\begin{equation}
  \ln\left(W_y^{(r)}\right)=r\ln(y)+\ln(W_y)
\end{equation}
and then taking the derivative with respect to $y$
\begin{equation}
  \frac{\partial \ln(W_y^{(r)})}{\partial y} = \frac{r}{y } + \frac{\partial \ln(W_y)}{\partial y}
  \ .
\end{equation}
Keep in mind that $y=1,2,3,\cdots$ so for large $y$, an approximation can be made $r/y\approx 0$ so that
\begin{equation}
  \frac{\partial \ln(W_y^{(r)})}{\partial y} \approx \frac{\partial \ln(W_y)}{\partial y}
  \ .
\end{equation}
Therefore the maximum of $W_y^{(r)}$ is located at $y_\text{max}$ for all $r=0,1,2,\cdots$. The same method of evaluating $W_y$ can be used to evaluate $W_y^{(r)}$. For a given $r=0,1,2,\cdots$, the limit of the sum $\sum_{y=y_\text{l}}^{y_\text{u}} W^{(r)}_y$ are chosen such that $W_{y_\text{l}}$ and $W_{y_\text{u}}$ are less than $\epsilon W_{y_\text{max}}^{(r)}$ where $\epsilon$ is some small constant. The limits will be different for different values of $r$.

>>>>

The compound Poisson distribution is studied here because of the compound Poission-like behaviour from the detection of photons \citep{whiting2002signal, elbakri2003efficient, whiting2006properties}. It is defined by defining a latent variable $Y\sim\poisson(\lambda)$ with probability mass function (p.m.f.) $\prob(Y=y)=\euler^{-\lambda}\frac{\lambda^y}{y!}$ for $y=0,1,2,\dotsc$, where $\lambda>0$ is the Poisson rate parameter. Let $U_i$ be some independent and identically distributed (i.i.d.) latent random variables with probability density function (p.d.f.) $p_U(u)$ for $i=1,2,3,\dotsc$. Let $X$ be a compound Poisson random variable where
\begin{equation}
  X|Y = \sum_{i=1}^{Y}U_i \ .
  \label{eq:compoundPoisson_X|Y}
\end{equation}
The p.d.f.~of $X$ can be obtained by marginalising the joint p.d.f.
\begin{equation}
  p_X(x)=\sum_{y=0}^\infty p_{X|Y}(x|y)\prob(Y=y) \quad\text{for }x\geqslant 0
  \ .
\end{equation}
It should be noted that $X=0$ if and only if $Y=0$ and this happens with probability $\prob(Y=0)=\euler^{-\lambda}$. This implies that $X$ has probability mass at zero and probability density for positive numbers which results in the p.d.f.
\begin{equation}
  p_X(x) = 
  \begin{cases}
    \delta(x) \euler^{-\lambda}  & \text{ for } x=0 \\ 
    \sum_{y=1}^\infty p_{X|Y}(x|y)\euler^{-\lambda}\frac{\lambda^y}{y!} \quad\text{for } & \text{ for } x>0
  \end{cases}
\end{equation}
where $\delta(x)$ is the Dirac delta function.

@@

In an x-ray tube, photons are emitted as a Poisson process and each photon has some random energy due to bremsstrahlung and characteristic radiation. This shares similarities to the compound Poisson. Let $Y$ be the number of photons emitted for some time exposure $\tau$, then $Y\sim\poisson(\lambda)$. Each photon is assumed to be i.i.d.~with energy $U_i$ for $i=1,2,3,\dotsc$. $U_i$ has p.d.f.~$p_U(u)$. The random variables discussed here covers all the latent variables in the compound Poisson.

@@

The evaluation of the p.d.f.~using the saddlepoint approximate required a bit of caution to avoid over/underflow problems. Suppose realizations of $X$ were simulated $\left\{x_1,x_2,x_3,\dotsc,x_n\right\}$. The saddlepoint approximation was computed up to a constant using
\begin{equation}
  p_X(x) \propto
  \exp\left[
    -\frac{\alpha+2}{2(\alpha+1)}
    \ln(x)
    -x\beta
    +\left(
    \frac{x\beta}{\alpha}
    \right)^{\frac{\alpha}{\alpha+1}}\lambda^{\frac{1}{\alpha+1}}(\alpha+1) - k
  \right]
\end{equation}
for 500 equally spaced points from and including the minimum non-zero simulated value to the maximum simulated value. $k$ is some constant which was chosen to be
\begin{equation}
  k =
  \max_{i\in\left\{1,2,3,\dotsc,n\right\}}
  \left[
    -\frac{\alpha+2}{2(\alpha+1)}
    \ln(x_i)
    -x_i\beta
    +\left(
      \frac{x_i\beta}{\alpha}
    \right)^{\frac{\alpha}{\alpha+1}}\lambda^{\frac{1}{\alpha+1}}(\alpha+1)
  \right]
  \ .
\end{equation}
The density was then was normalised by numerically integrating it using the trapezium rule using the 500 evaluated points.

A Q-Q plot is a plot which compares the empirical quantiles with the theoretical quantiles. Let
\begin{equation}
  F_X(x) = \prob(X\leqslant x)
\end{equation}
and
\begin{equation}
  \widehat{F}_X(x) = \frac{1}{n}\cdot\text{max}
  \left[
    \left(\sum_{i=1}^n\mathbb{I}(x_i\leqslant x)\right)-0.5,0
  \right]
  \ ,
\end{equation}
then a Q-Q plot is a parametric plot which plots $\widehat{F}_X^{-1}(p)$ against $F_X^{-1}(p)$ for $p=\frac{0.5}{n},\frac{1.5}{n},\frac{2.5}{n},\dotsc,\frac{n-0.5}{n}$. If $F_X(x)$ and $\widehat{F}_X(x)$ are similar, a Q-Q plot should be a straight line with gradient 1, intercepting the origin. For the exact method and the saddlepoint approximate, $F_X(x)$ was found numerically by evaluating the p.d.f.~at $10\,000$ equally spaced points from and including the minimum to the maximum of the simulated samples, and then summing the required trapeziums. $\widehat{F}_X^{-1}(p)$ was then calculated by interpolation. Because the Normal and saddlepoint approximation does not support zero, the Q-Q plot was omitted for small $\lambda$.

@@

The EM algorithm \citep{dempster1977maximum} is proposed as a method to estimate the parameters of a $X\sim\CPoisson(\lambda,\alpha,\beta)$ random variable given a sample of measurements of it $\left\{x_1,x_2,x_3,\dotsc,x_n\right\}$. The use of the EM algorithm here for the compound Poisson is similar to \cite{elbakri2003statistical, xu2009electronic}.

@@

The EM algorithm \citep{dempster1977maximum} instead treats $Y\sim\poisson(\lambda)$ as a latent variable. Let $\left\{Y_1,Y_2,Y_3,\dotsc, Y_n\right\}$ be realizations of $Y$ and $\left\{x_1,x_2,x_3,\dotsc, x_n\right\}$ be the respective measurements of $X|Y$. Define the joint log likelihood to be
\begin{multline*}
  \ln L(\lambda,\alpha,\beta;X,Y)
  =
  \sum_{i=1}^n
  \left[
    \mathbb{I}(x_i=0)
    \ln
    \prob(Y=0)
  \right.
  \\
  \left.
    +
    \mathbb{I}(x_i>0)
    \ln
    \left[
      p_{X|Y}(x_i|Y_i)\prob(Y=Y_i)
    \right]
  \right]
\end{multline*}
so that
\begin{multline}
  \ln L(\lambda,\alpha,\beta;X,Y)=
  \sum_{i=1}^n
  \left[
    -\mathbb{I}(x_i=0)
    \lambda
  \right.
  \\
  \left.+
    \mathbb{I}(x_i>0)
    \left(
      Y_i\alpha\ln\beta-\ln\Gamma(Y_i\alpha)+(Y_i\alpha-1)\ln x_i - \beta x_i
    \right.
  \right.
  \\
  \left.
    \left.  
      - \lambda + Y_i \ln \lambda - \ln(Y_i!)
    \right)
  \right]
  \ .
\end{multline}
The estimators are found by optimising the joint log likelihood. This is done iteratively by estimating the $Y$s given the $X$s and parameters (E step), followed by estimating the parameters given the $Y$s and $X$s (M step) until some convergence conditions are met.

@@

Let
\begin{equation}
  W_y^{(r)} = y^r W_y \quad \text{for }r=1,2,3,\dotsc
\end{equation}
so that
\begin{equation}
  y_i = \frac{\sum_{y=1}^\infty W^{(1)}_y}{\sum_{y=1}^\infty W_y} \ .
\end{equation}
Similarly
\begin{equation}
  \zeta_i = \variance[Y|X=x_i] = \frac{\sum_{y=1}^\infty W^{(2)}_y}{\sum_{y=1}^\infty W_y} - \left(y_i\right)^2
  \ .
\end{equation}
The expectation terms, $y_i$ and $\zeta_i$, are evaluated in the M step so that they can be used in the E step. The evaluation can be done by truncating the sum
\begin{equation}
  \sum_{y=1}^\infty W^{(r)}_y \approx \sum_{y=y_\text{l}}^{y_\text{u}} W^{(r)}_y
\end{equation}
where $y_\text{l}<y_\text{max}<y_\text{u}$ and $y_\text{max}$ is the value of $y$ which maximises $W_y^{(r)}$. By expressing $\ln(W^{(r)}_y)$ as
\begin{equation}
  \ln\left(W_y^{(r)}\right)=r\ln(y)+\ln(W_y)
\end{equation}
and then taking the derivative with respect to $y$
\begin{equation}
  \frac{\partial \ln(W_y^{(r)})}{\partial y} = \frac{r}{y } + \frac{\partial \ln(W_y)}{\partial y}
  \ .
\end{equation}
Keep in mind that $y=1,2,3,\dotsc$ so for large $y$, an approximation can be made $r/y\approx 0$ so that
\begin{equation}
  \frac{\partial \ln(W_y^{(r)})}{\partial y} \approx \frac{\partial \ln(W_y)}{\partial y}
  \ .
\end{equation}
Therefore the maximum of $W_y^{(r)}$ is located at $y_\text{max}$ for all $r=0,1,2,\dotsc$. The same method of evaluating $W_y$ can be used to evaluate $W_y^{(r)}$. For a given $r=0,1,2,\dotsc$, the limit of the sum $\sum_{y=y_\text{l}}^{y_\text{u}} W^{(r)}_y$ are chosen such that $W_{y_\text{l}}$ and $W_{y_\text{u}}$ are less than $\epsilon W_{y_\text{max}}^{(r)}$ where $\epsilon$ is some small constant. The limits will be different for different values of $r$.

### 2019-08-08 reports/thesis/chapters/4_variancePrediction.tex
inference chapter
subroi for more projections
use of dotsc
plot sub roi section

The sample variance-mean data were obtained from the replicated projections. Let $x_{i,j}^{(r)}$ be the grey value of the pixel at $(i,j)$ from the $r$th replicate projection for $r=1,2,3,\dotdotdot,n$. The pixel sample mean is
\begin{equation}
    \overline{x}_{i,j}=\frac{1}{n}\sum_{r=1}^n x_{i,j}^{(r)}
\end{equation}
and the pixel sample variance is
\begin{equation}
    y_{i,j} =
    \frac{1}{n-1}
    \sum_{r=1}^n
        \left(
            x_{i,j}^{(r)} - \overline{x}_{i,j}
        \right)^2
    \ .
\end{equation}
Only pixels in the region of interest (ROI) were considered here, that is pixels which represent the test sample. The ROI was created by manually segmenting the test sample from the projection.

The aim is to model and predict the grey value variance of a pixel given the pixel's grey value by fitting a model onto the sample variance-mean data. No spatial information was used, thus the variance-mean data is denoted as $(x_1,y_1),(x_2,y_2),\cdots,(x_m,y_m)$ from now on, where $m$ is the area of the ROI or the size of the dataset. Let $Y(x)$ be a random variable and the sample variance given a grey value mean $x$. It was assumed the standard error from estimating the mean was negligible so that the uncertainty is captured by the random variable $Y$. It was assumed that for a given pixel, the grey values are Normal and i.i.d. Let $\sigma^2(x)$ be the variance given a grey value $x$, then it can be shown that
\begin{equation}
\dfrac{(n-1)Y(x)}{\sigma^2(x)}\sim\chi^2_{n-1}
\end{equation}
which results in
\begin{equation}
Y(x)\sim\gammaDist\left(\alpha,\dfrac{\alpha}{\sigma^2(x)}\right)
\end{equation}
where $\alpha=(n-1)/2$ is the shape parameter. The expectation and variance are
\begin{equation}
\expectation\left[Y(x)\right]=\sigma^2(x)
\end{equation}
and
\begin{equation}
\variance\left[Y(x)\right]=\dfrac{\sigma^2(x)}{\alpha}
\end{equation}
respectively.

@@

An example of a systematic component are polynomial features $\eta(x)=\beta_0+\sum_{j=1}^{p-1}\beta_j x^j$ so that when used with the identity link, for example, $\expectation\left[Y(x)\right] = \beta_0+\sum_{j=1}^{p-1}\beta_j x^j$. Iterative reweighted least squares \citep{friedman2001elements} can be used to estimate the parameters $\beta_0, \beta_1, \cdots, \beta_{p-1}$ given data for the model to fit onto.

@@

>>>>

The sample variance-mean data were obtained from the replicated projections. Let $x_{i,j}^{(r)}$ be the grey value of the pixel at $(i,j)$ from the $r$th replicate projection for $r=1,2,3,\dotsc,n$. The pixel sample mean is
\begin{equation}
    \overline{x}_{i,j}=\frac{1}{n}\sum_{r=1}^n x_{i,j}^{(r)}
\end{equation}
and the pixel sample variance is
\begin{equation}
    y_{i,j} =
    \frac{1}{n-1}
    \sum_{r=1}^n
        \left(
            x_{i,j}^{(r)} - \overline{x}_{i,j}
        \right)^2
    \ .
\end{equation}
Only pixels in the region of interest (ROI) were considered here, that is pixels which represent the test sample. The ROI was created by manually segmenting the test sample from the projection.

The aim is to model and predict the grey value variance of a pixel given the pixel's grey value by fitting a model onto the sample variance-mean data. No spatial information was used, thus the variance-mean data is denoted as $(x_1,y_1),(x_2,y_2),\dotsc,(x_m,y_m)$ from now on, where $m$ is the area of the ROI or the size of the dataset. Let $Y(x)$ be a random variable and the sample variance given a grey value mean $x$. It was assumed the standard error from estimating the mean was negligible so that the uncertainty is captured by the random variable $Y$. It was assumed that for a given pixel, the grey values are Normal and i.i.d. Let $\sigma^2(x)$ be the variance given a grey value $x$, then it can be shown that
\begin{equation}
\dfrac{(n-1)Y(x)}{\sigma^2(x)}\sim\chi^2_{n-1}
\end{equation}
which results in
\begin{equation}
Y(x)\sim\gammaDist\left(\alpha,\dfrac{\alpha}{\sigma^2(x)}\right)
\end{equation}
where $\alpha=(n-1)/2$ is the shape parameter. The expectation and variance are
\begin{equation}
\expectation\left[Y(x)\right]=\sigma^2(x)
\end{equation}
and
\begin{equation}
\variance\left[Y(x)\right]=\dfrac{\sigma^2(x)}{\alpha}
\end{equation}
respectively.

@@

An example of a systematic component are polynomial features $\eta(x)=\beta_0+\sum_{j=1}^{p-1}\beta_j x^j$ so that when used with the identity link, for example, $\expectation\left[Y(x)\right] = \beta_0+\sum_{j=1}^{p-1}\beta_j x^j$. Iterative reweighted least squares \citep{friedman2001elements} can be used to estimate the parameters $\beta_0, \beta_1, \dotsc, \beta_{p-1}$ given data for the model to fit onto.

@@

With the ability to predict the variance given a grey value of a pixel, the uncertainty can be quantified. In the next chapter, the projection is compared to a simulated projection under the face of that uncertainty.
